### 2016-03-10 paper/icfp16-submission/semantics.tex
add intuition text for 3.2 and clean up recursion para

Next, we formalize the notion of type error witnesses as follows.
%
First, we define a core calculus within which we will work~\S~\ref{sec:syntax}.
%
Second, we develop a (non-deterministic) operational semantics
for ill-typed programs that precisely defines the notion
of a \emph{witness}~\S~\ref{sec:semantics}.
%
Third, we formalize and prove a notion of \emph{generality} for
witnesses, which states, intuitively, that if we find a
single witness then for \emph{every possible} type
assignment there exist inputs that are guaranteed to make
the program ``go wrong''~\S~\ref{sec:soundness}.
%
Finally, we refine the operational semantics into a
\emph{search procedure} that returns concrete (general)
witnesses for ill-typed programs~\S~\ref{sec:search-algorithm}.

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers, booleans, and binary trees.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ state in our syntax.

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a
notion of a \emph{hole}, written \vhole{\thole}, which represents
an \emph{unconstrained} value $\ehole$ that may be replaced with
\emph{any} value of type \thole.
%
Intuitively, the type holes \thole\ can be viewed as type variables
that we will \emph{not} generalize over.
%
A \emph{normalized} value is one that is not a hole,
but which may internally contain holes.
%
For example
$\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$ is a
normalized value.

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
value (resp.\ type) \emph{substitutions} $\vsu$ (resp. $\tsu$)
that map value (resp.\ type) holes to values (resp.\ types).
%
A \emph{normalized} value substitution is one whose co-domain is
comprised of normalized values.
%
In the sequel, we will assume and ensure that all value substitutions
are normalized.

@@

\subsection{Semantics}
\label{sec:semantics}
%
\RJ{Add intuition text -- "dynamic Hindley Milner"}
%
The substitutions let us ensure that we consistently instantiate
each hole with the same (partially defined) value or type, regardless
of the multiple contexts in which the hole appears. This ensures we
can report a concrete (and general) witness for any (dynamically)
discovered type errors.

@@

\paragraph{Generating Values} The (non-deterministic)
$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
as input a type $t$ and returns a value of that type.
%
For base types, the procedure returns an arbitrary value of
that type.
%
For functions, it returns a lambda with a \emph{new} hole
denoting the return value, and unconstrained types (denoted
by $\thole$) yield fresh holes constrained to have type
\thole (denoted by $\vhole{\thole}$).
%
When generating a $\ttree{t}$ we must take care to ensure
the resulting tree is well-typed. For a polymorphic type
$\ttree{\thole}$, eg in \recasegoodone, we will
place holes in the generated nodes; they will be lazily
filled in later, on demand.

@@

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A program state is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$ of an
expression $e$ or the stuck term $\stuck$, a value substitution $\vsu$,
and a type substitution $\tsu$. In the sequel we will write \estuck to
denote either an expression $e$ or \stuck.
%
We write $\step{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if the state
$\triple{\estuck}{\vsu}{\tsu}$ transitions in a \emph{single step} to
$\triple{\estuck'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of configurations
$\triple{\estuck_0}{\vsu_0}{\tsu_0}, \ldots, \triple{\estuck_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{\estuck_i}{\vsu_i}{\tsu_i}{\estuck_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,\triple{\estuck'}{\vsu'}{\tsu'}$.
%
We write \steps{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if
\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} for some trace $\trace$.

\paragraph{Primitive Reductions}
%
\RJ{Put high-level intuition about how "dynamic HM" is
formalized in op-sem, to set up next few paragraphs}
%
Primitive reduction steps --- addition, if-elimination,
function application, and data construction and case
analysis --- use \forcesym to ensure that values have
the appropriate type (and that holes are instantiated)
before continuing the computation.
%
Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that ``the caller'' $v_1$ is indeed
a function.

\paragraph{Recursion}
\RJ{this para appears out of nowhere --- non-sequitur}
Fixed-point operators often cannot be typed in static type
systems, but we are not concerned with \emph{assigning}
types to terms, rather with showing that \emph{no type}
can be assigned.
%
We are simply executing the untyped $\lambda$-calculus,
which has no issue handling recursion.

>>>>

Next, we formalize the notion of type error witnesses as follows.
%
First, we define a core calculus within which we will work~(\S~\ref{sec:syntax}).
%
Second, we develop a (non-deterministic) operational semantics
for ill-typed programs that precisely defines the notion
of a \emph{witness}~(\S~\ref{sec:semantics}).
%
Third, we formalize and prove a notion of \emph{generality} for
witnesses, which states, intuitively, that if we find a
single witness then for \emph{every possible} type
assignment there exist inputs that are guaranteed to make
the program ``go wrong''~(\S~\ref{sec:soundness}).
%
Finally, we refine the operational semantics into a
\emph{search procedure} that returns concrete (general)
witnesses for ill-typed programs~\S~(\ref{sec:search-algorithm}).

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers, booleans, and binary trees.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ term in our syntax. We write $\estuck$ to
denote terms that may be \stuck, and $e$ to denote terms that may not be
stuck.

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a
notion of a \emph{hole}, written \vhole{\thole}, which represents
an \emph{unconstrained} value $\ehole$ that may be replaced with
\emph{any} value of an unknown type \thole.
%
Intuitively, the type holes \thole\ can be viewed as type variables
that we will \emph{not} generalize over.
%
A \emph{normalized} value is one that is not a hole,
but which may internally contain holes.
%
For example
$\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$ is a
normalized value.

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
value (resp.\ type) \emph{substitutions} $\vsu$ (resp. $\tsu$)
that map value (resp.\ type) holes to values (resp.\ types).
%
The substitutions let us ensure that we consistently instantiate each
hole with the same (partially defined) value or type, regardless of the
multiple contexts in which the hole appears.
%
This ensures we can report a concrete (and general) witness for any
(dynamically) discovered type errors.

A \emph{normalized} value substitution is one whose co-domain is
comprised of normalized values.
%
In the sequel, we will assume and ensure that all value substitutions
are normalized.

@@

\subsection{Semantics}
\label{sec:semantics}
%
Recall that our goal is to synthesize a value that demonstrates
why (and how) a function goes wrong.
%
We accomplish this by combining evaluation with type inference,
giving us a form of dynamic type inference.
% starting with an unconstrained hole and gradually refining it
% as evaluation constrains its type.
%
Each primitive evaluation step tells us more about the types of the
program values. For example, addition tells us that the addends must be
integers, and case-analysis tells us that the scrutinee must be a tree.
%
When a hole appears in such a context, we know what type it must have
in order to make progress and can fill it in with a concrete value.

@@

\paragraph{Generating Values} The (non-deterministic)
$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
as input a type $t$ and returns a value of that type.
%
For base types, the procedure returns an arbitrary value of
that type.
%
For functions, it returns a lambda with a \emph{new} hole
denoting the return value, and unconstrained types (denoted
by $\thole$) yield fresh holes constrained to have type
\thole (denoted by $\vhole{\thole}$).
%
When generating a $\ttree{t}$ we must take care to ensure
the resulting tree is well-typed.
%
For a polymorphic type $\ttree{\thole}$ %, eg in \recasegoodone,
we will place holes in the generated nodes; they will be lazily filled
in later, on demand.

@@

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A configuration is a triple $\triple{\estuck}{\vsu}{\tsu}$ of an
expression $e$ or the stuck term $\stuck$, a value substitution $\vsu$,
and a type substitution $\tsu$.
%
% We write \estuck to
% denote either an expression $e$ or \stuck.
%
We write $\step{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if the state
$\triple{\estuck}{\vsu}{\tsu}$ transitions in a \emph{single step} to
$\triple{\estuck'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of configurations
$\triple{\estuck_0}{\vsu_0}{\tsu_0}, \ldots, \triple{\estuck_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{\estuck_i}{\vsu_i}{\tsu_i}{\estuck_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,\triple{\estuck'}{\vsu'}{\tsu'}$.
%
We write \steps{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if
\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} for some trace $\trace$.

\paragraph{Primitive Reductions}
%
% \RJ{Put high-level intuition about how "dynamic HM" is
% formalized in op-sem, to set up next few paragraphs}
%
Primitive reduction steps --- addition, if-elimination,
function application, and data construction and case
analysis --- use \forcesym to ensure that values have
the appropriate type (and that holes are instantiated)
before continuing the computation.
%
Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that ``the caller'' $v_1$ is indeed
a function.

\paragraph{Recursion}
%
% \RJ{this para appears out of nowhere --- non-sequitur}
Our semantics lacks a built-in $\mathtt{fix}$ construct
for defining recursive function, which may surprise
the reader.
%
Fixed-point operators often cannot be typed in static type
systems, but our system would simply approximate its type
as $\tfun$, apply it, and move along with evaluation.
%
Thus we do not need a built-in recursion construct, we
can use any of the standard fixed-point operators.

%  but we are not concerned with \emph{assigning}
% types to terms, rather with showing that \emph{no type}
% can be assigned.
% %
% We are simply executing the untyped $\lambda$-calculus,
% which has no issue handling recursion.
